---
title: "STAT331 Final Project"
author: "David Ballester, Dylan Le, Nathan Spiess, Tyler Stoen"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Load packages
```{r}
library(tidyverse)
library(here)
```

## Github operations
```{r}
library(usethis)
library(gitcreds)
use_git_config(user.name = "dle79", user.email = "dle79@calpoly.edu")
create_github_token()
gitcreds_set()
git_sitrep()
```

## Read in data
```{r}
age <- read.csv(here::here("final-project", "median_age_years.csv"))
health <- read.csv(here::here("final-project", "total_health_spending_per_person_us.csv"))
```
We will be using the two quantitative variables Total health spending per person in USD and median age.

Our response variable will be Total health spending / person USD which can be described as the average amount spent per person in USD for every country from 1994 to 2009 according to the World Health Organization.

Our observational variable will be the median age of the total population in each country since 1949. It also includes predicted median age until 2099.

We expect the relationship between average spending per person and median age to be positive because as the population becomes increasingly older on average, people will require more medical attention. We would also expect both the total health spending per person and the median age to increase with time.

For our age dataset, there was no cleaning required as there were no missing values or outlying values. All that was required for this data set was a pivot and then a merge with the health data set. 

## Pivot and Clean Data
```{r}
age_long <- age |>
  pivot_longer(cols = X1949:X2099,
               names_to = "year",
               values_to = "median_age") |>
  mutate(year = as.integer(str_sub(year, 2)))

health_long <- health |>
  pivot_longer(cols = X1994:X2009,
               names_to = "year",
               values_to = "healthCost") |>
  mutate(year = as.integer(str_sub(year, 2)))
```
Because the variable name started with a number, "X" was added before each year automatically by R. As a consequence, the year was changed into a categorical variable. Thus, after pivoting, I removed the "X" from each year and converted it into an integer.

Otherwise, no other data needed to be cleaned here.

## Join data
```{r}
ageHealth <- health_long |>
  left_join(age_long, by = c("country", "year"))
```
We decided to perform a left join between health_long and age_long, since the total health spending has less observation, and in turn, decrease the amount of missing values.





## Fitting a Simple Linear Model (Part 2.2)
```{r}
model <- lm(healthCost ~ median_age,
            data = ageHealth)

```

```{r}
library(broom)
tidy(model)
augment(model) |>
  ggplot(aes(x = median_age, y = .resid)) +
  geom_point() +
  geom_jitter()
```
