---
title: "Effect of Median Age on Total Health Spending"
author: "David Ballester, Dylan Le, Nathan Spiess, Tyler Stoen"
format: 
  html:
    self-contained: true
    code-tools: true
    toc: true
    number-sections: true
    code-fold: true
editor: source
execute: 
  error: true
  echo: true
  message: false
  warning: false
---

## Load packages
```{r}
library(tidyverse)
library(here)
```

```{r}
age <- read.csv(here::here("stat331-final-project", "median_age_years.csv"))
health <- read.csv(here::here("stat331-final-project", "total_health_spending_per_person_us.csv"))
```
We will be using the two quantitative variables Total health spending per person in USD and median age, all of which came from Gapminder.

Our response variable will be Total health spending / person USD which can be described as the average amount spent per person in USD for every country from 1994 to 2009 according to the World Health Organization.

Our observational variable will be the median age of the total population in each country since 1949. It also includes predicted median age until 2099.

We expect the relationship between average spending per person and median age to be positive because as the population becomes increasingly older on average, people will require more medical attention. We would also expect both the total health spending per person and the median age to increase with time.


Some cleaning needed to be done in order to make the data easier to work with.

## Pivot and Clean Data
```{r}
age_long <- age |>
  pivot_longer(cols = X1949:X2099,
               names_to = "year",
               values_to = "median_age") |>
  mutate(year = as.integer(str_sub(year, 2)))

health_long <- health |>
  pivot_longer(cols = X1994:X2009,
               names_to = "year",
               values_to = "healthCost") |>
  mutate(year = as.integer(str_sub(year, 2)))
```
Because the variable name started with a number, "X" was added before each year automatically by R. As a consequence, the year was changed into a categorical variable. Thus, after pivoting, I removed the "X" from each year and converted it into an integer.

Otherwise, no other data needed to be cleaned here.

```{r}
ageHealth <- health_long |>
  left_join(age_long, by = c("country", "year"))
```
We decided to perform a left join between health_long and age_long, since the total health spending has less observation, and in turn, decrease the amount of missing values.

```{r}
ageHealth |>
  ggplot(aes(y=healthCost,
             x=median_age)) +
  geom_jitter(width = 0.25, height = 0.5) +
  labs(title = "Total Health Spending vs. Median Age From 1994-2009",
       subtitle = "Health Spending",
       x = "Median Age",
       y = "") +
  geom_smooth(method = "lm") +
  theme_bw()
```
Based on the scatter plot above, our assumption was correct and there does appear to be a positive correlation between amount spent on health and median age. As the median age increases, the average amount spent on health does as well. Although the correlation appears to be positive, it does not appear linear.

```{r}
ageHealth |>
  ggplot(aes(y=healthCost,
             x=median_age,
             color=year)) +
  geom_jitter(width = 0.25, height = 0.5) + 
  labs(title = "Total Health Spending vs. Median Age From 1994-2009",
       subtitle = "Health Spending",
       x = "Median Age",
       y = "",
       color = "Year:") +
  theme_bw()
```
After coloring by year, we can see that within each of the shades of blue, the correlation appears to be pretty similar, so the relationship between amount spent on health and median age has not changed much form 1994-2009. 

Note: Although the data do not seem to be following a linear relationship based on the data visualizations above, we will use a linear relationship for the purposes of the project.

```{r}
model <- lm(healthCost ~ median_age,
            data = ageHealth)

```
We are fitting a linear relationship using median age to predict total health spending per person. So the x variable is median age and the y variable is total health spending.


```{r}
library(broom)
tidy(model)
```
Model:
predicted total health spending / person = -1745 + 95.4(median age)

For every one year increase in median age, there is a 95.4 USD increase in expected total health spending per person. A country with median age 0 has an expected total health spending per person of -1745 USD.

```{r}
library(moderndive)
var <- model |>
  get_regression_points() |>
  summarize(SSTotal = var(healthCost),
            SSModel = var(healthCost_hat),
            SSError = var(residual))

knitr::kable(var, "html",
             caption = "Variance of Model",
              col.names = c("SSTotal", "SSModel", "SSError"),
             )
```
In the table, we can see that 604462.2/1490797 * 100 = 40.5% of the variation in Health Costs is explained by the median age of the country. The R^2^ value is moderately large, which indicates our model has an adequate fit when comparing total healthcare costs and median age.
